---
title: "genRelateR Package Tutorial"
author: "Kobi Jamal Schmalenberg"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{genRelateR_tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction
`genRelateR` genRelateR is a tool for analyzing global human genomic relatedness across populations. It enhances workflows in bioinformatics by providing exploratory analysis tools for genomic data. The package includes the main components: DESCRIPTION, NAMESPACE, man subdirectory and R subdirectory. Additionally, licence, README and subdirectories vignettes, tests, and data are also available. It was written in R Markdown, using the [knitr](https://cran.r-project.org/package=knitr) package for production. See `help(package = "genRelateR")` for further details. 

To download **genRelateR**, use the following commands:

``` r
require("devtools")
devtools::install_github("bdls-jamal/genRelateR", build_vignettes = TRUE)
library("genRelateR")
```
To list all sample functions available in the package:
``` r
ls("package:genRelateR")
```

To list all sample datasets available in the package:
``` r
data(package = "genRelateR")
```
You may add more datasets from https://www.internationalgenome.org/data/

## Initialization and Environment Setup
```{r setup, include = FALSE}
library(genRelateR) # Load the package
```
Run the file: R/setupPackages.R, to install and load necessary packages.

Ensure that lines 131 and 135 in loadAndCleanData.R show the correct file path
to population code and long/latitude information text files.

## Running the functions
### Downloading VCF Data
``` r
# Data in example is from the 1000 genomes project. This package is for the project and may not work as intended with other data that does not have the same format as the project.
# Define the base FTP URL(you may choose a different release)
base_url <- "ftp://ftp.1000genomes.ebi.ac.uk/vol1/ftp/release/20130502/"

# Specify the filenames for the VCF and .tbi files(example only below)
vcf_file <- "ALL.chr1.phase3_shapeit2_mvncall_integrated_v5b.20130502.genotypes.vcf.gz"
tbi_file <- paste0(vcf_file, ".tbi")

# Specify the destination folder (create it if it doesn’t exist)
dest_folder <- "data/vcf_files"
if (!dir.exists(dest_folder)) {
  dir.create(dest_folder, recursive = TRUE)
}

# Define the full destination paths
vcf_dest <- file.path(dest_folder, vcf_file)
tbi_dest <- file.path(dest_folder, tbi_file)

# Download the VCF and .tbi files
download.file(url = paste0(base_url, vcf_file), destfile = vcf_dest, mode = "wb")
download.file(url = paste0(base_url, tbi_file), destfile = tbi_dest, mode = "wb")

# Download population metadata files, or grab from github repo
pop_codes_url <- "https://raw.githubusercontent.com/bdls-jamal/genRelateR/main/inst/extdata/population_codes.txt"
pop_long_lat_url <- "https://raw.githubusercontent.com/bdls-jamal/genRelateR/main/inst/extdata/population_long_lat.txt"
pop_metadata_url <- "https://raw.githubusercontent.com/bdls-jamal/genRelateR/main/inst/extdata/population_metadata.txt"

dest_folder <- "data"

pop_codes_dest <- file.path(dest_folder, "population_codes.txt")
pop_long_lat_dest <- file.path(dest_folder, "population_long_lat.txt")
pop_metadata_dest <- file.path(dest_folder, "population_metadata.txt")

download.file(url = pop_codes_url, destfile = pop_codes_dest, mode = "wb")
download.file(url = pop_long_lat_url, destfile = pop_long_lat_dest, mode = "wb")
download.file(url = pop_metadata_url, destfile = pop_metadata_dest, mode = "wb")

```
### Functional Calling
``` r
Run genRelateR::setupGeneticPackages() to install and load necessary packages

You may download the necessary txt files directly from the github repository or from the 1000 genomes project along with the vcf.gz/tbi files from the project.

# Load genetic data from compressed vcf file(i.e. data from 1000 genomes project)
# Ensure this data has an associated tbi file in the same folder, or generate it using Tabix
# Example: vcf_file <- "inst/extdata/vcf/ALL.chr1.phase3_shapeit2_mvncall_integrated_v5b.20130502.genotypes.vcf.gz"
vcf_file <- "path/to/your/vcf.gz/data"

# Define a region(range) for analysis(required for very large data files)
chr1_region <- GRanges(
  seqnames = "1",
  ranges = IRanges(start = 20000000, end = 20001000)
)

# Call loadGeneticData function with specified region and vcf file
genetic_data <- genRelateR::loadGeneticData(vcf_file, regions = chr1_region)

# Create population metadata object from metadata file
# File is provided for 1000 genome project use, for other data you must use with your own data
# Example: pop_metadata <- "inst/extdata/population_metadata.txt"
pop_metadata <- "path/to/your/metadata.txt/file"

# Choose populations to filter by(all are selected in this example)
populations <- c(
  "CHB", "JPT", "CHS", "CDX", "KHV",
  "CEU", "TSI", "GBR", "FIN", "IBS",
  "YRI", "LWK", "GWD", "MSL", "ESN",
  "ASW", "ACB", "MXL", "PUR", "CLM",
  "PEL", "GIH", "PJL", "BEB", "STU", "ITU"
)

# Call filterPopulation with loaded data, metadata, and specified populations
filtered_data <- genRelateR::filterPopulation(genetic_data, pop_metadata, populations)

# Perform population structure analysis by calling analyzePopulationStructure
pca_results <- genRelateR::analyzePopulationStructure(
  filtered_data$vcf_data,
  filtered_data$pop_metadata,
  method = "pca"
)

# Create PCA data cluster visualization
# Call plotPopulationPca then print to view
pca_plot <- genRelateR::plotPopulationPca(
  analysis_results = pca_results,
  filtered_data$pop_metadata,
  title = "Population Structure PCA",
  ellipses = TRUE,
  super_pop = TRUE
)
print(pca_plot)

# Create PCA data map visualization
# Call plotAncestryMap then print to view
ancestry_map <- genRelateR::plotAncestryMap(
  pca_results,
  filtered_data$pop_metadata,
  title = "Global Population Distribution"
)
print(ancestry_map)

```

## Package References

Schmalenberg, K. (2024) genRelateR: R Package Development for BCB410H. Unpublished. https://github.com/bdls-jamal/genRelateR

## Other References

----
```{r}
sessionInfo()
```
